Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM trees)

### 2015-08-25 pkg/glmertree/paper/treatment_interactions paper.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@755 255f4249-2263-4417-b1ea-8498779e27a7

\usetikzlibrary{shapes,shadows,arrows,positioning}
\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM tree)}
\shorttitle{GLMM Tree}
\leftheader{GLMM Tree}
\rightheader{GLMM Tree}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

@@

%\section{Citation stuff for journal publication}
%
%Potential Psychological Methods references: only \citeA{StroyMall09} (not added yet). Maybe force in: 
%
%''Hedges, L. V., \& Pigott, T. D. (2004). The power of statistical tests for moderators in meta-analysis. 
%
%''Cooper, H., \& Patall, E. A. (2009). The relative benefits of meta-analysis conducted with individual participant data versus aggregated data. Psychological methods, 14(2), 165.'' (they write in abstract ''Regardless of whether a meta-analysis is conducted with IPD or AD, synthesists must remain vigilant in how they interpret their results. They must avoid ecological fallacies, Simpson’s paradox, and interpretation of synthesis-generated evidence as supporting causal inferences.'')
%
%''Hedges, L. V., \& Vevea, J. L. (1998). Fixed-and random-effects models in meta-analysis. Psychological methods, 3(4), 486.''

@@

In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Personalized medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

@@

Before we discuss these methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated. After we introduce the GLMM tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy. Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.

@@

To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment 1 (78 observations) or Treatment 2 (72 observations). Every observation has a value on the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. Further, all observations have values on three covariates: duration of depressive symptoms prior to treatment in months (range 0-15); age in years (range 18-75); anxiety inventory total score (range 3-18).

The simulated dataset has 3 subgroups with treatment interactions. The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$. In this subgroup, the mean of the response variable for Treatment 1 is 7, and the mean for the response variable for Treatment 2 is 11. The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$. In this subgroup, the mean value of the response variable for Treatment 1 and 2 is 9. The third subgroup consists of observations with duration $>6$. In this subgroup, the mean value of the response variable for Treatment 1 is 12, and the mean of the response variable for Treatment 2 is 7.

@@

Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function. In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}$. Further, $x_{i}^{\top}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.

To keep notation and examples simple, we assume $x_{i}^{\top}$ and $\beta$ to have length 2. That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the GLM. However, additional treatment conditions and covariates can easily be included. In addition, examples and datasets in the current paper will focus on continuous response variables with normally distributed errors, such as posttreatment severity of a disorder. But the models and algorithms to be discussed can also be applied with discrete outcomes, such as remission of a disorder (yes/no).

@@

Alternatively, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM. The model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*\top}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{* \top}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

@@

For simplicity, we assume $z_{i}$ and $b_{m}$ to have length 1 in the current paper; that is, only cluster-specific intercepts are included in the models. However, random-effects covariates and coefficients can easily be included. Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b_{m}$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the linear predictor of a GLM.

@@

To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12}. \citeA{HajjyBell11} and \citeA{SelaySimo12} developed a method for estimation of mixed-effects regression trees (MERTs), which are somewhat similar to GLMM trees. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

For estimating GLMM trees, we take the MERT approach a step further, by using a GLM tree instead of a regression tree with constant fits. This allows not only for detection of differences in main effects,but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM-tree algorithm takes the following steps to estimate the model in Equation \ref{eq:glimmertree}:

@@

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_{k},U_{k'}}=0.0$, $\rho_{U_{k},U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42). 
\item Two different levels for $\beta_1$, the unstandardized mean difference in treatment outcomes, in subgroups with differential effects for Treatment 1 ($x_{1}=0$) and Treatment 2 ($x_{1}=1$). The levels for mean differences in subgroups with differential treatment effect were $|\beta_1| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_1| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

@@

@@

To generate node-specific fixed effects, we partitioned the sample according to the terminal nodes of the tree in Figure 4.3. In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$. For every observation, we generated a binomial variable (with $p =.5$) as an indicator for treatment type.

@@

To detect predictors of tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The effects of predictors with main and/or interaction effects with $\eta^2 > .01$ were further investigated using graphical displays.

To detect predictors of tree accuracy for both algorithms in datasets with treatment-subgroup, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The effects of predictors with main and/or interaction effects with unstandardized coefficients $|\beta| > .2$ were further investigated using graphical displays.

@@

For every dataset, two correlation coefficients were calculated, representing the linear association between the true and predicted treatment-effect differences: one for GLM tree, and one for GLMM tree. To detect predictors of predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The effects of predictors with main and/or interaction effects with $\eta^2 > .01$ were further investigated using graphical displays.

@@

\subsubsection{Estimation of GLM and GLMM trees}

@@

For estimating GLMM trees, we implemented the GLMM tree algorithm in a function that iterates between (1) estimation of a generalized linear model tree using the \verb|lmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|lmer| function. Convergence of \verb|glmmtree| is monitored using the log-likelihood value of the generalized linear mixed-effects model estimated in step (2) of the algorithm. When the difference in the log-likelihoods of two consecutive iterations is less than a prespecified value (.001 by default), \verb|glmmtree| has converged.

For building the GLM trees in step (1) of the GLMM tree algorithm, the same settings as described above were used. That is, models in the nodes of the tree were estimated with OLS, the $\alpha$-level for assessing parameter instability was set to .05, the Bonferroni correction for multiple testing was applied, the minimum number of observations in a node was set to 20, and maximum tree depth was set to four.

@@

\subsection{Tree size and accuracy in datasets without treatment-subgroup interactions}

@@

\begin{table}[!htbp]
\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
\small
\begin{tabular}{lrrrrrrr}
	\thickline
	&\multicolumn{7}{c}{tree size}\\
	\cline{2-7}
	&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
	\hline
	GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
		  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
   	GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
    			& (.67)  & (.20) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
	\hline
  	\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
  	\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
  	\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
\end{tabular}
\label{tab:treesize_main}
\end{table}

\begin{figure}[!htbp]
	\includegraphics[width=12cm]{xy_treesizes_maineff.pdf}
	\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
	\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\subsection{Tree size in datasets with treatment-subgroup interactions}

In datasets with treatment-subgroup interactions, GLMM trees were also smaller than GLM trees. For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure \ref{fig:modelC}). Th distribution of tree sizes for GLM and GLMM tree in datasets with treatment-subgroup interactions are presented in Table \ref{tab:treesize_interactions}. The average size of GLMM trees was 7.15 (SD=0.61), and the average size of GLM trees was 8.11 (SD=2.05). The estimated probability that a datasets was erroneously not partitioned was 0, for both GLM and GLMM tree. However, Table \ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .64 of GLM trees matched the true tree size (Table \ref{tab:treesize_interactions}).

\begin{table}[!htbp]
\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
\small
\begin{tabular}{lrrrrrrrr}
	\thickline
	&&\multicolumn{7}{c}{tree size}\\
	\cline{2-8}
	&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
	\hline
  	GLMM tree	& 3 & 227& 29556 & 2472 & 89 & 3 &0& 32,400 \\
				&  ($<.01$)  & ($<.01$)& (.91) & ($<.01$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\

  	GLM tree		& 145 & 1002 & 20578 & 4443 & 4178 & 1665 & 389 & 32,400\\ 
 			 	& ($<.01$) & (.03) & (.64) & (.14) & (.13) & (.05) & (.01) & (1.00) \\ 
	\hline
  	\multicolumn{9}{l}{\textit{Note. } Bracketed values are proportions. Tree sizes are expressed as the total number of}\\
  	\multicolumn{9}{l}{nodes in the tree. A tree with a total of $J$ nodes has $(J+1)/2$ terminal nodes; the true}\\	
  	\multicolumn{8}{l}{tree size in datasets with treatment-subgroup interactions was 7.}\\
\end{tabular}
\label{tab:treesize_interactions}
\end{table}

A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure \ref{fig:xyplot_treesize_interact}). When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees of size 7, on average.

Clear differences in performance between GLM and GLMMtree were observed when $\sigma_b > 0$. When $b$ is not correlated with one of the $U_k$ variables, when sample size is small (i.e., 200) and when $\sigma_b$ is large (i.e., 10), GLM tree has difficulty detecting splits and grows trees that are too small, on average. When $b$ is not correlated with one of the $U_k$ variables and when sample size is larger (i.e., 500 or 1000), GLM and GLMM trees are about the same size (i.e., $\approx 7$). When $b$ is correlated with one of the $U_k$ variables, GLM starts creating spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10).

\begin{figure}[!htbp]
	\includegraphics[width=12cm]{xy_treesizes_treatsubs.pdf}
	\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=7$ represents true tree size.}
	\label{fig:xyplot_treesize_interact}
\end{figure}

\subsection{Tree accuracy in datasets with treatment-subgroup interactions}

To assess the accuracy of the trees created by GLM and GLMM tree, we inspected the variables and values that were selected for partitioning in every dataset. For the first split, both GLM tree and GLMM tree always selected the true partitioning variable ($U_2$). The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split was 29.94, for both GLM and GLMM tree. However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split, than did GLMM tree (SD=0.155 and SD=0.126, respectively).

@@

A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the probability of accurate tree recovery for both algorithms (Figure \ref{fig:xyplot_treeaccuracy}). When random effects were absent from the datasets (i.e., $\sigma_b = 0$), the trees recovered by GLM and GLMM tree were equally accurate, on average. In the presence of random effects, GLM trees were much less accurate than GLMM trees. This was found for all sample sizes, when $b$ was correlated to one of the $U_k$ variables. When $b$ was not correlated to one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., 200).

\begin{figure}[!htbp]
	\includegraphics[width=12cm]{xy_accuracy.pdf}
	\caption{Average accuracy of GLM and GLMM trees. Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered. Values ``correlated'' and ``uncorrelated'' refer to whether random intercept values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
	\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

To assess predictive accuracy of both algorithms, correlation between the true and predicted treatment-effect differences of both algorithms were calculated for every dataset. Overall, treatment-effect differences predicted by GLMM tree were closer to the true differences than those predicted by GLM tree. The average correlation between the true and predicted treatment-effect differences over all 32,400 datasets was .88 (SD=0.20) for GLM tree, and .94 (SD=0.10) for GLMM tree.

>>>>

\usetikzlibrary{shapes,shadows,arrows,positioning}
\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM trees)}
\shorttitle{GLMM Tree}
\leftheader{GLMM Tree}
\rightheader{GLMM Tree}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

@@

@@

In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and stratified medicine is becoming increasingly important. Stratified medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

@@

Before we discuss these methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated. After we introduce the GLMM-tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy. Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.

@@

To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment 1 or Treatment 2. Every observation has a value on the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. Further, all observations have values on three covariates: duration of depressive symptoms prior to treatment in months (range 0-15); age in years (range 18-75); anxiety inventory total score (range 3-18).

The simulated dataset has 3 subgroups with different treatment effectiveness. The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$. In this subgroup, Treatment 1 is more effective than Treatment 2: the mean of the response variable is 7 for Treatment 1, and 11 for Treatment 2. The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$. In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment 1, and 9 for Treatment 2. The third subgroup consists of observations with duration $>6$. In this subgroup, Treatment 2 is more effective than Treatment 1: the mean value of the response variable is 12 for Treatment 1, and 7 for Treatment 2.

@@

Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function. In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}^2$. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (taking a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type). $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.

To keep notation and examples simple, we assume $x_{i}$ and $\beta$ to have length 2. That is, the effects of only two treatment conditions are estimated and no additional covariates are included in the GLM. However, additional treatment conditions and covariates can easily be included. In addition, examples and datasets in the current paper will focus on continuous response variables with normally distributed errors, such as posttreatment severity of a disorder. But the models and algorithms to be discussed can also be applied with discrete outcomes, such as remission of a disorder (yes/no).

@@

Alternatively, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM. The model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

@@

For simplicity, we assume $z_{i}$ and $b_{m}$ to have length 1 in the current paper; that is, only cluster-specific intercepts are included in the models. However, random-effects covariates and coefficients can easily be included. Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b_{m}$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the GLM.

@@

To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12}. They developed a method for the estimation of mixed-effects regression trees (MERTs), which are somewhat similar to GLMM trees. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

For estimating GLMM trees, we take the MERT approach a step further: by using a GLM tree, instead of a CART regression tree with constant fits, to estimate the fixed-effects part of the GLMM. This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM-tree algorithm takes the following steps to estimate the model in Equation \ref{eq:glimmertree}:

@@

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_{k},U_{k'}}=0.0$, $\rho_{U_{k},U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42). 
\item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects. The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

@@

It should be noted that when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.

@@

To generate node-specific fixed effects, we partitioned the sample according to the terminal nodes of the tree in Figure 4.3. In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$. For every observation, we generated a binomial variable (with probability .5) as an indicator for treatment type.

@@

To assess the impact of data-generating parameters on tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.

To assess the impact of data-generating parameters on tree accuracy in datasets with treatment-subgroup interactions, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The impact of predictors with main and/or interaction effects with unstandardized regression coefficients $> .5$ (i.e., an in- or decrease in the log-odds of .5) were further investigated using graphical displays.

@@

For every dataset, correlation coefficients for each algorithm were calculated, representing the linear association between the true and predicted treatment-effect differences. To assess the impact of data-generating parameters on predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.

@@

\subsubsection{Estimation of GLM trees}

@@

For estimating GLMM trees, we implemented the GLMM-tree algorithm in a function that iterates between (1) estimation of a generalized linear model tree using the \verb|lmtree| function, and (2) estimation of the random-effects coefficients $b_m$ using the \verb|lmer| function. Convergence of \verb|glmmtree| is monitored using the log-likelihood value of the generalized linear mixed-effects model estimated in step (2) of the algorithm. When the difference in the log-likelihoods of two consecutive iterations is less than a prespecified value (.001 by default), \verb|glmmtree| has converged.

For building the GLM trees in step (1) of the GLMM-tree algorithm, the same settings as described above were used. That is, models in the nodes of the tree were estimated with OLS, the $\alpha$-level for assessing parameter instability was set to .05, the Bonferroni correction for multiple testing was applied, the minimum number of observations in a node was set to 20, and maximum tree depth was set to four.

@@

\subsubsection{Tree size in datasets without treatment-subgroup interactions}

@@

\begin{table}[!htbp]
\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
\small
\begin{tabular}{lrrrrrrr}
	\thickline
	&\multicolumn{7}{c}{tree size}\\
	\cline{2-7}
	&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
	\hline
	GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
		  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
   	GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
    			& (.67)  & (.19) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
	\hline
  	\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
  	\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
  	\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
\end{tabular}
\label{tab:treesize_main}
\end{table}

\begin{figure}[!htbp]
	\includegraphics[width=15cm]{xy_treesizes_maineff.pdf}
	\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
	\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\subsubsection{Tree size in datasets with treatment-subgroup interactions}

In datasets with treatment-subgroup interactions, GLMM trees were also smaller than GLM trees. For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure \ref{fig:modelC}). Th distribution of tree sizes for GLM and GLMM tree in datasets with treatment-subgroup interactions are presented in Table \ref{tab:treesize_interactions}. The average size of GLMM trees was 7.16 (SD=0.62), and the average size of GLM trees was 8.12 (SD=2.05). The estimated probability that a dataset was erroneously not partitioned was 0, for both GLM and GLMM tree. However, Table \ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .63 of GLM trees matched the true tree size (Table \ref{tab:treesize_interactions}).

\begin{table}[!htbp]
\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
\small
\begin{tabular}{lrrrrrrrr}
	\thickline
	&&\multicolumn{7}{c}{tree size}\\
	\cline{2-8}
	&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
	\hline
  	GLMM tree	& 2 & 281& 29358 & 2675 & 83 & 1 &0& 32,400 \\
				&  ($<.01$)  & ($<.01$)& (.91) & ($.08$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\

  	GLM tree		& 139 & 955 & 20530 & 4565 & 4164 & 1619 & 428 & 32,400\\ 
 			 	& ($<.01$) & (.03) & (.63) & (.14) & (.13) & (.05) & (.01) & (1.00) \\ 
	\hline
  	\multicolumn{9}{l}{\textit{Note. } Bracketed values are proportions. Tree sizes are expressed as the total number of}\\
  	\multicolumn{9}{l}{nodes in the tree. A tree with a total of $J$ nodes has $(J+1)/2$ terminal nodes; the true}\\	
  	\multicolumn{8}{l}{tree size in datasets with treatment-subgroup interactions was 7.}\\
\end{tabular}
\label{tab:treesize_interactions}
\end{table}

A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure \ref{fig:xyplot_treesize_interact}). When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees with a size of about 7, on average.

\begin{figure}[!htbp]
	\includegraphics[width=15cm]{xy_treesizes_treatsubs.pdf}
	\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=7$ represents true tree size.}
	\label{fig:xyplot_treesize_interact}
\end{figure}

Clear differences in performance between GLM and GLMMtree were observed when $\sigma_b > 0$. When $b$ is not correlated with one of the $U_k$ variables, when sample size is small (i.e., 200) and when $\sigma_b$ is large (i.e., 10), GLM tree has difficulty detecting splits and grows trees that are too small, on average. When $b$ is not correlated with one of the $U_k$ variables and when sample size is larger (i.e., 500 or 1000), GLM and GLMM trees are about the same size (i.e., $\approx 7$). When $b$ is correlated with one of the $U_k$ variables, GLM creates spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10). This effect was stronger when $b$ was correlated to a non-splitting variable.

\subsubsection{Tree accuracy in datasets with treatment-subgroup interactions}

To assess the accuracy of the trees created by GLM and GLMM tree, we inspected the variables and values that were selected for partitioning in every dataset. For the first split, GLMM tree always selected the true partitioning variable ($U_2$). GLM tree selected a wrong partitioning variable (I.e., $U_1$ in only one dataset. The true splitting value for $U_2$ was 30 (Figure \ref{fig:modelC}), and the mean splitting value selected for the first split (involving $U_2$) was 29.94, for both GLM and GLMM tree. However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split (involving $U_2$), than did GLMM tree (SD=0.154 and SD=0.127, respectively).

@@

A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on the probability of accurate tree recovery for both algorithms (Figure \ref{fig:xyplot_treeaccuracy}). When random effects were absent from the datasets (i.e., $\sigma_b = 0$), the trees recovered by GLM and GLMM tree were equally accurate, on average. In the presence of random effects, GLM trees were much less accurate than GLMM trees. This was found for all sample sizes, when $b$ was correlated to one of the $U_k$ variables, and the effect was somewhat stronger when the correlated $U_k$ was not a true partitioning variable. When $b$ was not correlated to one of the $U_k$ variables, GLMM tree clearly outperformed GLM tree only when sample size was small (i.e., 200).

\begin{figure}[!htbp]
	\includegraphics[width=15cm]{xy_accuracy.pdf}
	\caption{Average accuracy of GLM and GLMM trees. Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered. Values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
	\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

To assess predictive accuracy of both algorithms, correlation between the true and predicted treatment-effect differences of both algorithms were calculated for every dataset. Overall, treatment-effect differences predicted by GLMM tree were closer to the true differences than those predicted by GLM tree. The average correlation between the true and predicted treatment-effect differences over all 32,400 datasets was .88 (SD=0.19) for GLM tree, and .94 (SD=0.11) for GLMM tree.
